---
title: "Loris_indicator_taxa_analysis"
output: html_document
---

## Differential Abundance Analysis

For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

Since my work on the diversity analysis took less time than expected, I will also work on Differential abundance analysis, to fulfill Aim 2 of the final research paper. Due to the indicator species analysis being pretty much already completed, I'm using this script to investigate differential abundance with ANCOM-BC2 instead. I will do two sepparate analysis, one where I aggregate to genus level and one where I aggregate to family or class level. I may use this file to also do core microbiome analysis as well, if time permits and no-one has worked on it yet.

#### Differential abundance with ANCOM-BC2 (genus-level):

Here is my code for the differential abundance analysis. Anything notable about the results will be mentioned directly in the code chunks as blocked out text.

```{r Indicator Taxa Analysis, fig.height=9, fig.width=8.5}
library(tidyverse)
library(phyloseq)
library(microbiome)
library(indicspecies)
library(ANCOMBC)

phyloseq = readRDS('../Datasets/phyloseq_taxonomy.rds')

# Aggregating to the genus level
ps_genus_glom = tax_glom(phyloseq, 'Genus')

set.seed(421)
ancombc2_out = ancombc2(data = ps_genus_glom, 
               fix_formula = 'Time_of_Collection_categorical*dose_categorical', 
               p_adj_method = 'BH',
               prv_cut = 0.1)

statistical_table = ancombc2_out$res

#writexl::write_xlsx(statistical_table,'../Results/Tables/Module 14 - ANCOMBC Results.xlsx')

# Filter stats to only include taxa that are differentially abundant between the time points:
taxa_to_plot = statistical_table %>% 
  filter(diff_robust_Time_of_Collection_categorical30_days==T) 

# Alternatively, since the previous method got lots of falses (only 1 paired diff robust collumn returned true):
ancombc2_by_dose <- function(dose_for_ancom) {
  # Subsetting the phyloseq object (like in beta diversity)
  ps_sub <- subset_samples(ps_genus_glom,
                           dose_categorical == get(dose_for_ancom))
  # Droping taxa with zero counts after the subsetting
  ps_sub <- prune_taxa(taxa_sums(ps_sub) > 0, ps_sub)
  # Run ANCOM-BC2 test
  out <- ancombc2(
    data = ps_sub,
    fix_formula = "Time_of_Collection_categorical",
    p_adj_method = "BH",
    prv_cut = 0.1
  )
  # Extracting the results
  res <- out$res %>%
    mutate(dose = dose_for_ancom)

  return(res)
}

# function 2.0 (runs but takes a terribly long time):
ancombc2_by_dose <- function(dose_for_ancom) {

  samples_keep <- sample_names(ps_genus_glom)[
    sample_data(ps_genus_glom)$dose_categorical == dose_for_ancom
  ]

  ps_sub <- prune_samples(samples_keep, ps_genus_glom)

  ps_sub <- prune_taxa(taxa_sums(ps_sub) > 0, ps_sub)

  out <- ancombc2(
    data = ps_sub,
    fix_formula = "Time_of_Collection_categorical",
    p_adj_method = "BH",
    prv_cut = 0.1
  )

  res <- out$res %>%
    mutate(dose = dose_for_ancom)

  return(res)
}

dose_levels = unique(sample_data(ps_genus_glom)$dose_categorical)

all_DA_results = lapply(dose_levels, ancombc2_by_dose) %>%
  bind_rows()

# Plot the log 2 fold changes:
genus_table = tax_table(ps_genus_glom) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("taxon")

all_DA_results = all_DA_results %>%
  left_join(genus_table, by = "taxon")

all_DA_results %>%
  ggplot(aes(Family,lfc_Time_of_Collection_categorical30_days, fill = Class)) +
  geom_col() +
  coord_flip() +
  theme_classic(base_size = 13) +
  facet_grid(~dose, scales = 'free_y') +
  facet_wrap(~dose, scales = 'free_y') +
  ylab('Log-2 fold change') + xlab(NULL) +
  labs(title = "Differential abundance for 4 radiation doses") +
  theme(legend.position = "bottom", legend.direction = "horizontal")

```

#### Differential abundance with ANCOM-BC2 (class-level):

This is basically the same thing as above but aggregated to a higher taxonomic level, to see if the plots look better. A new code chunk is being made since the entire analysis needs a new custom function.

```{r Class-level_analysis, fig.height=9, fig.width=8.5}
# the same phyloseq object is being used as above.
# aggregating to the class level:
ps_class_glom = tax_glom(phyloseq, 'Class')

# same function but tailored to the new phyloseq objects:
ancombc2_by_dose <- function(dose_for_ancom) {

  samples_keep <- sample_names(ps_class_glom)[
    sample_data(ps_class_glom)$dose_categorical == dose_for_ancom
  ]

  ps_sub <- prune_samples(samples_keep, ps_class_glom)

  ps_sub <- prune_taxa(taxa_sums(ps_sub) > 0, ps_sub)

  out <- ancombc2(
    data = ps_sub,
    fix_formula = "Time_of_Collection_categorical",
    p_adj_method = "BH",
    prv_cut = 0.1
  )

  res <- out$res %>%
    mutate(dose = dose_for_ancom)

  return(res)
}

dose_levels = unique(sample_data(ps_class_glom)$dose_categorical)

all_DA_results = lapply(dose_levels, ancombc2_by_dose) %>%
  bind_rows()

# plotting log2 fold changes:
class_table = tax_table(ps_class_glom) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("taxon")

all_DA_results = all_DA_results %>%
  left_join(class_table, by = "taxon")

all_DA_results %>%
  ggplot(aes(Class,lfc_Time_of_Collection_categorical30_days, fill = Phylum)) +
  geom_col() +
  coord_flip() +
  theme_classic(base_size = 13) +
  facet_grid(~dose, scales = 'free_y') +
  facet_wrap(~dose, scales = 'free_y') +
  ylab('Log-2 fold change') + xlab(NULL) +
  labs(title = "Differential abundance for 4 radiation doses") +
  theme(legend.position = "bottom", legend.direction = "horizontal")

# Visually it looks a lot better aggregated to class, not so squished as before. Even when plotting the genus-aggregation at family level, the whole thing was still super packed.
```

#### Differential abundance with ANCOM-BC2 (family level):

This is basically the same thing as above but aggregated to the Family level, to hopefully get the ideal plot. A new code chunk is being made since the entire analysis also needs a new custom function. Hopefully this will be less cramped than the genus aggregation but have better discernment than class or phylum aggregations.

```{r Family-level analysis, fig.height=9, fig.width=8.5}
ps_family_glom = tax_glom(phyloseq, 'Family')

# same function but tailored to the new phyloseq objects:
ancombc2_by_dose <- function(dose_for_ancom) {

  samples_keep <- sample_names(ps_family_glom)[
    sample_data(ps_family_glom)$dose_categorical == dose_for_ancom
  ]

  ps_sub <- prune_samples(samples_keep, ps_family_glom)

  ps_sub <- prune_taxa(taxa_sums(ps_sub) > 0, ps_sub)

  out <- ancombc2(
    data = ps_sub,
    fix_formula = "Time_of_Collection_categorical",
    p_adj_method = "BH",
    prv_cut = 0.1
  )

  res <- out$res %>%
    mutate(dose = dose_for_ancom)

  return(res)
}

dose_levels = unique(sample_data(ps_family_glom)$dose_categorical)

all_DA_results = lapply(dose_levels, ancombc2_by_dose) %>%
  bind_rows()

# plotting log2 fold changes:
family_table = tax_table(ps_family_glom) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("taxon")

all_DA_results = all_DA_results %>%
  left_join(family_table, by = "taxon")

all_DA_results %>%
  ggplot(aes(Family,lfc_Time_of_Collection_categorical30_days, fill = Phylum)) +
  geom_col() +
  coord_flip() +
  theme_classic(base_size = 13) +
  facet_grid(~dose, scales = 'free_y') +
  facet_wrap(~dose, scales = 'free_y') +
  ylab('Log-2 fold change') + xlab(NULL) +
  labs(title = "Differential abundance for 4 radiation doses") +
  theme(legend.position = "bottom", legend.direction = "horizontal")
```
