---
title: "Ramayani_core_microbiome_analysis"
output: html_document
date: "2026-03-20"
---

```{r}
library(tidyverse)
library(phyloseq)
library(tidyverse)
library(dplyr)
library(microbiome)
library(ggVennDiagram)

# Load object
ps = readRDS('../Datasets/phyloseq_taxonomy.rds')

ps_rare = ps %>% 
  rarefy_even_depth(sample.size = 87521, rngseed = 123)
```




```{r}

subset_samples(ps_rare, dose_categorical %in% c("0.1_dose", "0.25_dose", "1_dose"))
ps_transform = transform(ps_rare, 'compositional')
ps_species = tax_glom(ps_transform, 'Species')

# Core Microbiome Analyses for 10 and 30 days

# Subset your phyloseq object
ten_days_post_radiation = subset_samples(ps_species, Time_of_Collection_categorical == '10_days')
thirty_days_post_radiation = subset_samples(ps_species, Time_of_Collection_categorical == '30_days')

# Find core members
ASVs_10_days = core_members(ten_days_post_radiation, detection=0, prevalence = 0.8)
ASVs_30_days = core_members(thirty_days_post_radiation, detection=0, prevalence = 0.8)


venn_diagram = ggVennDiagram(list(ASVs_10_days, ASVs_30_days),
              set_size = 5,
              category.names = c('10_days','30_days'),
              label_size = 6, label_alpha = 0) + scale_fill_distiller(palette = "RdPu") +
  scale_x_continuous(expand = expansion(mult = .5)) + 
  labs(title = "Venn Diagram Depicting Count of Same/Different Microbiome Species
       Present Within the Gut 10 Days and 30 Days Post Radiation Exposure")
                                                       
venn_diagram

ggsave("../Results/RP_core_microbiome.png")