---
title: "differential abundance 2nd attempt"
output: html_document
date: "2026-03-27"
---

```{r}
library(Maaslin2)
library(tidyverse)
library(phyloseq)
library(microViz)
library(dplyr)
library(tibble)
library(stringr)
library(RColorBrewer)
library(ggplot2)
library(knitr)
library(dplyr)

set.seed(321)

ps = readRDS('../Datasets/phyloseq_taxonomy.rds')

ps_glom = tax_glom(ps, 'Genus')
```

Since need to subset the metadata into each dose group, ran the analysis on each dose seperately and plotted only significant genera as relative abundance plots.

```{r}
# 0 Gy

#subsetting phyloseq object for only 0 Gy
ps_0dose = subset_samples(ps_glom, dose_categorical %in% c("0_dose"))

df_abundance_0dose = as.data.frame(otu_table(ps_0dose))
if (taxa_are_rows(ps_0dose)) { df_abundance_0dose <- t(df_abundance_0dose) }

df_metadata_0dose = data.frame(sample_data(ps_0dose))

df_abundance_0dose = df_abundance_0dose[match(rownames(df_metadata_0dose), rownames(df_abundance_0dose)),]

# Attaching genus names to ASVs to get accurate information
tax_table_df_0dose = as.data.frame(tax_table(ps_0dose))

genus_names_0dose = as.character(tax_table_df_0dose$Genus)

# Wrangling of data to ensure correct pairing of ASVs and relevant genera
genus_names_0dose = stringr::str_remove(genus_names_0dose, " g__")

genus_names_0dose[is.na(genus_names_0dose) | genus_names_0dose == ""] = "Unclassified"

genus_names_0dose = make.unique(genus_names_0dose)

colnames(df_abundance_0dose) <- genus_names_0dose

# Running differential analysis using MaAsLin2 for 0 Gy
fit_data_0dose = Maaslin2(
    input_data = df_abundance_0dose,
    input_metadata = df_metadata_0dose,
    output = 'to_delete',
    fixed_effects = c("Time_of_Collection_categorical"),
    reference = 'Time_of_Collection_categorical,10_days',
    normalization = "TSS",
    transform = "AST",
    min_abundance = 0.001, 
  min_prevalence = 0.1, 
  max_significance = 0.05,
  plot_heatmap = FALSE,
  plot_scatter = FALSE
)

# Saving results as a table and excel file
statistical_table_0dose = fit_data_0dose$results

writexl::write_xlsx(statistical_table_0dose,'../Results/Maaslin2_Results_0dose.xlsx')
```


```{r}

# Filtering for significant genera

maaslin_output_0dose = statistical_table_0dose %>% mutate(log2fc = log2(exp(statistical_table_0dose$coef)))

taxa_to_plot_0dose = maaslin_output_0dose %>% filter(qval < 0.05, abs(log2fc) >= -1)

#Plotting results based on log2 fold change
taxa_to_plot_0dose %>%
  ggplot(aes(feature,coef)) +
  geom_col() +
  coord_flip()

taxa_to_plot_0dose

```


Since no significant genera found, no relative abundance plot created for 0 dose group.

```{r}
# 0.1 Gy


# subsetting phyloseq object for only 0.1 Gy
ps_0.1dose = subset_samples(ps_glom, dose_categorical %in% c("0.1_dose"))

df_abundance_0.1dose = as.data.frame(otu_table(ps_0.1dose))

if (taxa_are_rows(ps_0.1dose)) { df_abundance_0.1dose <- t(df_abundance_0.1dose) }

df_metadata_0.1dose = data.frame(sample_data(ps_0.1dose))
df_abundance_0.1dose = df_abundance_0.1dose[match(rownames(df_metadata_0.1dose), rownames(df_abundance_0.1dose)),]


# Attaching genus names to ASVs to get accurate information
tax_table_df_0.1dose = as.data.frame(tax_table(ps_0.1dose))

genus_names_0.1dose = as.character(tax_table_df_0.1dose$Genus)

# Wrangling of data to ensure correct pairing of ASVs and relevant genera
genus_names_0.1dose = stringr::str_remove(genus_names_0.1dose, " g__")

genus_names_0.1dose[is.na(genus_names_0.1dose) | genus_names_0.1dose == ""] = "Unclassified"

genus_names_0.1dose = make.unique(genus_names_0.1dose)

colnames(df_abundance_0.1dose) <- genus_names_0.1dose

# Running differential analysis using MaAsLin2 for 0.1 Gy
fit_data_0.1dose = Maaslin2(
    input_data = df_abundance_0.1dose,
    input_metadata = df_metadata_0.1dose,
    output = 'to_delete',
    fixed_effects = c("Time_of_Collection_categorical"),
    reference = 'Time_of_Collection_categorical,10_days',
    normalization = "TSS",
    transform = "AST",
    min_abundance = 0.001, 
  min_prevalence = 0.1, 
  max_significance = 0.05,
  plot_heatmap = FALSE,
  plot_scatter = FALSE
)

# Saving results as a table and excel file
statistical_table_0.1dose = fit_data_0.1dose$results

writexl::write_xlsx(statistical_table_0.1dose,'../Results/Maaslin2_Results_0.1dose.xlsx')
```


```{r}
# Filtering for significant genera

maaslin_output_0.1dose = statistical_table_0.1dose %>% mutate(log2fc = log2(exp(statistical_table_0.1dose$coef)))

taxa_to_plot_0.1dose = maaslin_output_0.1dose %>% filter(qval < 0.05, abs(log2fc) >= -1)

#Plotting results based on log2 fold change

taxa_to_plot_0.1dose %>%
  ggplot(aes(feature,log2fc, fill = feature)) +
  geom_col(color = "black", width = 0.7) +           # Clean bar outlines
  scale_fill_brewer(palette = "Set3") +             # Professional palette
  theme_classic(base_size = 14) +                   # Clean theme, larger font
  labs(
    title = "Log2 Fold Change of Differential Abundance at 30 days Post Radiation Exposure (0.1Gy)",,
    x = "Log2 Fold Change",
    y = "Genus"
  ) +
  theme(
    legend.position = "none",                       # Remove redundant legend
    plot.title = element_text(face = "bold"),       # Bold title
    axis.title = element_text(face = "plain")
  ) +
  coord_flip()

taxa_to_plot_0.1dose

```
Since significant genera for 0.1 Gy, will make relevant abundance plots down below.

```{r}
# 0.25 Gy

# subsetting phyloseq object for only 0.25 Gy
ps_0.25dose = subset_samples(ps_glom, dose_categorical %in% c("0.25_dose"))

df_abundance_0.25dose = as.data.frame(otu_table(ps_0.25dose))
if (taxa_are_rows(ps_0.25dose)) { df_abundance_0.25dose <- t(df_abundance_0.25dose) }

df_metadata_0.25dose = data.frame(sample_data(ps_0.25dose))

df_abundance_0.25dose = df_abundance_0.25dose[match(rownames(df_metadata_0.25dose), rownames(df_abundance_0.25dose)),]

# Attaching genus names to ASVs to get accurate information
tax_table_df_0.25dose = as.data.frame(tax_table(ps_0.25dose))

genus_names_0.25dose = as.character(tax_table_df_0.25dose$Genus)


# Wrangling of data to ensure correct pairing of ASVs and relevant genera
genus_names_0.25dose = stringr::str_remove(genus_names_0.25dose, " g__")

genus_names_0.25dose[is.na(genus_names_0.25dose) | genus_names_0.25dose == ""] = "Unclassified"

genus_names_0.25dose = make.unique(genus_names_0.25dose)

colnames(df_abundance_0.25dose) <- genus_names_0.25dose

# Running differential analysis using MaAsLin2 for 0.25 Gy
fit_data_0.25dose = Maaslin2(
    input_data = df_abundance_0.25dose,
    input_metadata = df_metadata_0.25dose,
    output = 'to_delete',
    fixed_effects = c("Time_of_Collection_categorical"),
    reference = 'Time_of_Collection_categorical,10_days',
    normalization = "TSS",
    transform = "AST",
    min_abundance = 0.001, 
  min_prevalence = 0.1, 
  max_significance = 0.05,
  # These plots increase run time and are usually poorly formatted, so you might as well stop them from being generated
  plot_heatmap = FALSE,
  plot_scatter = FALSE
)

# Saving results as a table and excel file
statistical_table_0.25dose = fit_data_0.25dose$results

writexl::write_xlsx(statistical_table_0.25dose,'../Results/Maaslin2_Results_0.25dose.xlsx')
```



```{r}
# Filtering for significant genera
maaslin_output_0.25dose = statistical_table_0.25dose %>% mutate(log2fc = log2(exp(statistical_table_0.25dose$coef)))

taxa_to_plot_0.25dose = maaslin_output_0.25dose %>% filter(qval < 0.05, abs(log2fc) >= -1)

#Plotting results based on log2 fold change
taxa_to_plot_0.25dose %>%
  ggplot(aes(feature,log2fc)) +
  geom_col() +
  coord_flip()

taxa_to_plot_0.25dose
```
Since no significant genera found, no relative abundance plot created for 0.25 Gy.

```{r}
# 1 Gy

# subsetting phyloseq object for only 1 Gy
ps_1dose = subset_samples(ps_glom, dose_categorical %in% c("1_dose"))

df_abundance_1dose = as.data.frame(otu_table(ps_1dose))
if (taxa_are_rows(ps_1dose)) { df_abundance_1dose <- t(df_abundance_1dose) }

df_metadata_1dose = data.frame(sample_data(ps_1dose))

df_abundance_1dose = df_abundance_1dose[match(rownames(df_metadata_1dose), rownames(df_abundance_1dose)),]

# Attaching genus names to ASVs to get accurate information
tax_table_df_1dose = as.data.frame(tax_table(ps_1dose))

genus_names_1dose = as.character(tax_table_df_1dose$Genus)

# Wrangling of data to ensure correct pairing of ASVs and relevant genera
genus_names_1dose = stringr::str_remove(genus_names_1dose, " g__")

genus_names_1dose[is.na(genus_names_1dose) | genus_names_1dose == ""] = "Unclassified"

genus_names_1dose = make.unique(genus_names_1dose)

colnames(df_abundance_1dose) <- genus_names_1dose

# Running differential analysis using MaAsLin2 for 1 Gy
fit_data_1dose = Maaslin2(
    input_data = df_abundance_1dose,
    input_metadata = df_metadata_1dose,
    output = 'to_delete',
    fixed_effects = c("Time_of_Collection_categorical"),
    reference = 'Time_of_Collection_categorical,10_days',
    normalization = "TSS",
    transform = "AST",
    min_abundance = 0.001, 
  min_prevalence = 0.1, 
  max_significance = 0.05,
  plot_heatmap = FALSE,
  plot_scatter = FALSE
)

# Saving results as a table and excel file
statistical_table_1dose = fit_data_1dose$results

writexl::write_xlsx(statistical_table_1dose,'../Results/Maaslin2_Results_1dose.xlsx')
```


```{r}
# Filtering for significant genera
maaslin_output_1dose = statistical_table_1dose %>% mutate(log2fc = log2(exp(statistical_table_1dose$coef)))

taxa_to_plot_1dose = maaslin_output_1dose %>% filter(qval < 0.05, abs(log2fc) >= -1)

#Plotting results based on log2 fold change
taxa_to_plot_1dose %>%
  ggplot(aes(feature,log2fc, fill = feature)) +
  geom_col(color = "black", width = 0.7) +           # Clean bar outlines
  scale_fill_brewer(palette = "Set3") +             # Professional palette
  theme_classic(base_size = 14) +                   # Clean theme, larger font
  labs(
    title = "Log2 Fold Change of Differential Abundance at 30 days Post Radiation Exposure (1Gy)",
    x = "Log2 Fold Change",
    y = "Genus"
  ) +
  theme(
    legend.position = "none",                       # Remove redundant legend
    plot.title = element_text(face = "bold"),       # Bold title
    axis.title = element_text(face = "plain")
  ) +
  coord_flip()

taxa_to_plot_1dose
```
Since significant genera for 1 Gy, will make relevant abundance plots down below.

```{r}
# 0.1 Gy

# get significant genus
taxa_to_plot_0.1dose = maaslin_output_0.1dose %>%
  filter(qval < 0.05, abs(log2fc) >= -1) %>%
  pull(feature)  %>% str_remove("g__") %>% 
  replace_na("Unclassified")

# find relative abundance
ps_rel_0.1dose = transform_sample_counts(ps_0.1dose,
                                         function(x) x / sum(x))
# Melt taxa
df_plot_0.1dose = psmelt(ps_rel_0.1dose)

# Clean genus names
df_plot_0.1dose$Genus = stringr::str_remove(df_plot_0.1dose$Genus, "g__")
df_plot_0.1dose$Genus = stringr::str_trim(df_plot_0.1dose$Genus)
df_plot_0.1dose$Genus[df_plot_0.1dose$Genus == "" | is.na(df_plot_0.1dose$Genus)] = "Unclassified"
df_plot_feature_0.1dose = df_plot_0.1dose %>%
  pull(Genus)

head(maaslin_output_0.1dose$feature)
head(df_plot_0.1dose$Genus)

# Keep only significant genera
df_plot_0.1dose = df_plot_0.1dose %>%
  filter(Genus %in% taxa_to_plot_0.1dose)

df_plot_0.1dose$Genus = stringr::str_replace_all(df_plot_0.1dose$Genus, "_", " ")
df_plot_0.1dose$Time_of_Collection_categorical = stringr::str_replace_all(df_plot_0.1dose$Time_of_Collection_categorical, "_", " ") 


boxplot_0.1dose = ggplot(df_plot_0.1dose,
       aes(x = Time_of_Collection_categorical,
           y = Abundance,
           color = Time_of_Collection_categorical)) +
  
  geom_boxplot(aes(fill = Time_of_Collection_categorical),
               outlier.shape = NA,
               alpha = 0.3) +
  
  geom_jitter(width = 0.2, size = 2, alpha = 0.8) +
  
  facet_wrap(~ Genus, scales = "free_y") +
  
  theme_classic() +
  
  labs(title = "Relative Abundance of Significant Genera (0.1 Gy)",
       x = "Days Post Radiation Exposure",
       y = "Relative Abundance") +
  
  theme(strip.text = element_text(face = "italic")) +
  
  # Soft pink + purple
  scale_color_manual(values = c("10 days" = "#F4A7B9",
                                "30 days" = "#B57EDC")) +
  scale_fill_manual(values = c("10 days" = "#F4A7B9",
                               "30 days" = "#B57EDC"))
boxplot_0.1dose

ggsave("../Results/RP_differential_abundance_0.1dose.png")
```
```{r}
# get significant genus
taxa_to_plot_1dose = maaslin_output_1dose %>%
  filter(qval < 0.05, abs(log2fc) >= -1) %>%
  pull(feature)  %>% str_remove("g__") %>% 
  replace_na("Unclassified")

# find relative abundance
ps_rel_1dose = transform_sample_counts(ps_1dose,
                                         function(x) x / sum(x))
# Melt taxa
df_plot_1dose = psmelt(ps_rel_1dose)

# Clean genus names
df_plot_1dose$Genus = stringr::str_remove(df_plot_1dose$Genus, "g__")
df_plot_1dose$Genus = stringr::str_trim(df_plot_1dose$Genus)
df_plot_1dose$Genus[df_plot_1dose$Genus == "" | is.na(df_plot_1dose$Genus)] = "Unclassified"
df_plot_feature_1dose = df_plot_1dose %>%
  pull(Genus)

head(maaslin_output_1dose$feature)
head(df_plot_1dose$Genus)

# Keep only significant genera
df_plot_1dose = df_plot_1dose %>%
  filter(Genus %in% taxa_to_plot_1dose)

df_plot_1dose$Genus = stringr::str_replace_all(df_plot_1dose$Genus, "_", " ")
df_plot_1dose$Time_of_Collection_categorical = stringr::str_replace_all(df_plot_1dose$Time_of_Collection_categorical, "_", " ") 


boxplot_1dose = ggplot(df_plot_1dose,
       aes(x = Time_of_Collection_categorical,
           y = Abundance,
           color = Time_of_Collection_categorical)) +
  
  geom_boxplot(aes(fill = Time_of_Collection_categorical),
               outlier.shape = NA,
               alpha = 0.3) +
  
  geom_jitter(width = 0.2, size = 2, alpha = 0.8) +
  
  facet_wrap(~ Genus, scales = "free_y", labeller = label_wrap_gen(width = 20)) +
  
  theme_classic() +
  
  labs(title = "Relative Abundance of Significant Genera (1 Gy)",
       x = "Days Post Radiation Exposure",
       y = "Relative Abundance") +
  
  theme(strip.text = element_text(face = "italic")) +
  
  # Soft pink + purple
  scale_color_manual(values = c("10 days" = "#F4A7B9",
                                "30 days" = "#B57EDC")) +
  scale_fill_manual(values = c("10 days" = "#F4A7B9",
                               "30 days" = "#B57EDC"))
boxplot_1dose

ggsave("../Results/RP_differential_abundance_1dose.png")
```

