---
title: "Loris_a_and_b_diversity_analysis"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

As my contribution to the term project for MICB 305, I will conduct the diversity analysis, both alpha and beta. I will format the plots to look publication appropriate, and the final plots I make will probably be the ones used in the manuscript, provided everything goes well. Feel free to let me know if there's anything I need to change, but try not to work on this script at the same time as I am, so we can avoid any issues with conflicting pushes!

#### Alpha diversity analysis:

Here is my code for the Alpha diversity analysis. Anything notable about the results will be mentioned directly in the code chunks as blocked out text.

```{r Alpha_diversity}
# summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
